---
title: "RNNs for Movie Review Sentiment Analysis"
author: " Quinlan Wilson, Akhil Gorla,Ella Yang, Lorreta Lu, Anish Senthil"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

In this vignette, we will build a Recurrent Neural Network (RNN) using an LSTM (Long Short-Term Memory) architecture to classify movie reviews as positive or negative.

RNNs are a powerful deep learning architecture designed to work with sequential data, such as text, speech, or time-series information. Unlike traditional neural networks, RNNs preserve information across time steps, allowing them to understand context, order, and dependencies.

Sentiment analysis is a classic application of natural language processing (NLP). In this example, we use an IMDB movie review dataset and train a model that learns to interpret word patterns and predict the sentiment of a review.

#### What is a RNN

A Recurrent Neural Network processes input one element at a time while maintaining a hidden "memory" state:

$$
h_t = \tanh \left( W_x x_t + W_h h_{t-1} + b \right)
$$

Where:

- \(x_t\) is the input at time step \(t\)
- \(h_t\) is the hidden state at time \(t\)
- \(h_{t-1}\) is the previous hidden state
- \(W_x, W_h\) are learned weight matrices
- \(b\) is a bias vector

LSTMs solve vanishing gradient problems by adding gates (input, forget, output) and maintaining a cell state.

## Dataset and Preprocessing

Before an RNN can learn from text, the text must be converted to numbers.  
We encode sentiment labels (positive = 1, negative = 0), tokenize text, restrict vocabulary to 10,000 words, and pad or truncate sequences to a fixed length.

```{r, warning=FALSE, message=FALSE}

# This vignette requires TensorFlow and Keras.
# If TensorFlow is not installed, run in the R console:
#   install.packages("tensorflow")
#   library(tensorflow)
#   install_tensorflow()
# This model is based on Keras
# If the Keras not installed or automatically installed Keras3
# Please change to Keras(Old version)
#   install.packages("keras)
#   library(keras)
#   install_keras()

library(keras) # for deep learning
library(tidyverse) # for read_csv and data manipulation
library(caret) # for train/test split
library(tensorflow)

set.seed(3434)
# Read in the data
movie_reviews <- read_csv("data/IMDB Dataset.csv")

# Select the column with info on the emotion of review
sentiment <- movie_reviews$sentiment

# Check class balance (should be about equal in our case exactly equal)
table(sentiment)

# Convert labels to binary (1 = positive, 0 = negative)
movie_reviews$sentiment_binary <- ifelse(movie_reviews$sentiment == "positive", 1, 0)

# Tokenization (converting text into integers)
# Neural networks can't work with raw text so tokenization is used.
# On a basic level tokenization changes each word into an integer based on frequency. 
# We only use the top 10k most frequent words
max_words <- 10000

# RNNs need all sequences to be the same length.
# So by choosing 500 if a review has less than 500 words
# those extras are removed and if it has less than 500
# we add zeros to make it the same length. 
max_len <- 250
review_len <- sapply(strsplit(movie_reviews$review, " "), length)
hist(review_len, breaks = 50, main="Distribution of Review Lengths", xlab = "Words per review", xlim = c(0, 1250))
summary(review_len)

# Create a tokenizer and fit it to our review text
tokenizer <- text_tokenizer(num_words = max_words)
tokenizer %>% fit_text_tokenizer(movie_reviews$review)

word_counts <- tokenizer$word_counts
word_freq <- sort(unlist(word_counts), decreasing = T)
# Top 20 words
head(word_freq, 20)

# How many words appear only once or twice
sum(word_freq == 1)
sum(word_freq == 2)

# Since only 52k words appear once and 15k twice setting
# max words to 10k makes sense as these rare words add
# much to our classification model.
# Convert each review to a sequence of integers
sequences <- texts_to_sequences(tokenizer, movie_reviews$review)

# Make every sequences have the same length
x <- pad_sequences(sequences = sequences, maxlen = max_len)
y <- movie_reviews$sentiment_binary

# split the dataset into training and testing (80/20)
train_index <- createDataPartition(y, p = 0.8, list = F)
x_train <- x[train_index, ]
x_test <- x[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]
```

The dataset contains IMDB movie reviews with text and sentiment labels *(positive/negative)*. We first convert the sentiment labels into binary values, mapping positive $\rightarrow$ 1 and negative $\rightarrow$ 0.

Next, we perform tokenization, which converts each word into an integer index based on its frequency or identity. For example:

- “the" $\rightarrow$ 666757

- "movie" $\rightarrow$ 87050

- "film" $\rightarrow$ 77678

This step transforms variable-length text into sequences of integers that the RNN can process.

Because reviews differ in length, we then pad or truncate each sequence to a fixed maximum length. Reviews longer than the limit are cut, and shorter ones are padded with zeros. This ensures a uniform input shape for the model.

Finally, we restrict the vocabulary to the top 10,000 most frequent words, removing extremely rare words that add noise and do not help with training.

## Model Architecture: Embedding + Bidirectional LSTM

Before sequences enter the LSTM, they pass through an embedding layer that learns a dense vector for each word.

The embedding captures semantic similarity; e.g., "good" is closer to "great" than "terrible".

We use 64-dimensional embedding.  
The output for each review is a matrix (sequence length $\times$ 64).

```{r}
embedding_dim <- 64

model <- keras_model_sequential() %>%
  layer_embedding(
    input_dim  = max_words,
    output_dim = embedding_dim

  ) %>%
  bidirectional(
    layer_lstm(units = 64, dropout = 0.2, recurrent_dropout = 0.2)
  ) %>%
  layer_dense(units = 1, activation = "sigmoid")
summary(model)
```
Model architecture:

1. **Embedding layer** - converts word indices to 64-dimensional vectors.  
2. **Bidirectional LSTM** - outputs a 128-dimensional representation.  
3. **Dense output layer** - sigmoid activation gives probability of positive sentiment.

This allows capturing word meaning, order, and context.

#### Training the Model

We train the model using:

- Adam optimizer  
- Binary cross-entropy loss  

Training involves multiple epochs, with a validation split to monitor overfitting.
```{r, eval=FALSE}
# Compile the model: Adam optimizer + binary cross-entropy for binary labels
model %>% compile(
  optimizer = "adam",
  loss = "binary_crossentropy",
  metrics = "accuracy"
)

# Train the model for 3 epochs.
# validation_split = 0.2 uses 20% of the training data as validation
history <- model %>% fit(
  x_train, y_train,
  epochs = 3,
  batch_size = 30,
  validation_split = 0.2
)

# Plot accuracy and loss over epochs (passes)
plot(history)

```
We compile the model with the Adam optimizer and binary cross-entropy loss, then train for 3 epochs while monitoring validation performance.

Training history shows changes in accuracy and loss. Ideally, both training and validation loss decrease or stabilize.

#### Visual Word Embeddings

We extract the learned embedding matrix and apply PCA to reduce 64 dimensions to 2 for visualization.

```{r, eval=FALSE}
# Extract the embedding matrix from the first layer
embeddings <- get_weights(model$layers[[1]])[[1]]

# Perform PCA on a subset of embeddings (skip index 1 which is usually padding)
pca_res <- prcomp(embeddings[2:1000, ])

# Take the first 2 principal components
pca_df <- data.frame(
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2],
  word = names(tokenizer$word_index)[1:999]
)

# Scatterplot of words in the 2D PCA space
ggplot(pca_df, aes(x = PC1, y = PC2, label=word)) +
  geom_text(size=3) +
  ggtitle("Word Embeddings PCA Projection")
```

We project the 64-dimensional word embeddings down to 2 dimensions using PCA. Words with similar sentiment (e.g., great, wonderful, favorite vs awful, worst, terrible) tend to cluster together, while neutral words lie near the center.

A scatterplot then reveals:

- positive sentiment words clustered together  
- negative sentiment words near each other  
- common neutral words in the center  

This helps us interpret learned semantic structures.

#### Evaluation

After training, we now evaluate the trained model on the held-out test set to assess how well it generalizes to unseen reviews.

```{r, eval=FALSE}
# Evaluate the model on the held-out test data
model %>% evaluate(x_test, y_test)
```

Metrics returned include accuracy and loss.  
A good model should achieve high accuracy and low loss on unseen data.

#### Predicting Sentiment on New Examples

We test the model on new sentences (tokenized and padded like training data).  
Predictions > 0.5 $\rightarrow$ positive.

Examples:

- "This is the best movie I have ever seen!" $\rightarrow$ positive  
- "The acting in this movie was horrible." $\rightarrow$ negative  

This shows generalization to new text.
```{r, eval=FALSE}
# Example sentences not seen during training
examples <- c(
  "This is the best movie I have ever seen!",
  "The acting in this movie was horrible."
)

# Tokenize and pad the example sentences in the same way as the training data

examples_seq <- texts_to_sequences(tokenizer, examples)
examples_pad <- pad_sequences(examples_seq, maxlen= max_len)

# Predict sentiment probabilities
predictions <- model %>% predict(examples_pad)

# Convert probabilities to class labels
pred_label <- ifelse(predictions > 0.5, "positive", "negative")
pred_label
```

#### Limitations on Model

LSTMs have limitations:

- Cannot handle misspellings or unseen words  
- Long documents may still lose information  
- No attention mechanism; all words weighted similarly  

More advanced models (GRUs, Transformers, BERT, GPT) outperform LSTMs on modern NLP tasks.

But LSTMs remain excellent for learning sequence modeling fundamentals.
